\\pkg{casebase}: An Alternative Framework for Survival Analysis and Comparison of Event Rates

### 2022-11-19 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd

Case-base sampling allows us to avoid this integral. By conditioning on a sampled person-moment, we get individual partial likelihood contributions of the form
$$P(\,\textrm dN_{i}(t) \mid \,\textrm dQ_{i}(t) = 1,\mathrm{past}) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{\,\textrm dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.$$
By using the symbol $\stackrel{\theta}{\propto}$, we mean that both sides of the expression are equal up to multiplicative factors that do not depend on $\theta$. Therefore, an estimating function for $\theta$ can be composed of these contributions as:
\begin{equation}
L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{\,\textrm dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{\,\textrm dQ_i(t)}. \label{eq:lik-function}
\end{equation}
When a logarithmic link function is used for modeling the hazard function, the above expression is of a logistic regression form with an offset term $\log(1/\rho_i(t))$. Note that the sampling units selected in the case-base sampling mechanism are person-moments, rather than individuals, and the parameters to be estimated are hazards or hazard ratios rather than odds or odds ratios. Generally, an individual can contribute more than one person-moment, and thus the terms in the product integral are not independent. Nonetheless, Saarela [-@saarela2016case] showed that the corresponding partial likelihood score function has mean zero at the true value $\theta=\theta_0$, and that the resulting estimator $\widehat{\theta}$ is asymptotically normally distributed.

@@

  1. Compute $\widehat{\theta}$ and its variance-covariance matrix $V(\widehat \theta)$ which is obtained as part of the logistic regression output.
  2. Sample $B$ times from a multivariate normal $N\left(\widehat{\theta}, V(\widehat\theta)\right)$.
  3. For each sample, compute the survival function using Equation \ref{eqn:surv} below.
  4. Use the pointwise quantiles of these survival function estimates to construct a pointwise confidence band for the survival function of interest.

@@

\begin{equation}
\log(\lambda(t; X)) = \beta_0 + \beta_1 X. \label{eq:exp}
\end{equation}
In this model, the instantaneous failure rate is independent of $t$.^[The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study. This is also known as the *memoryless property* [@kalbfleisch2011statistical].]

@@

\begin{equation}
\log(\lambda(t; X)) = \beta_0 + \beta_1 t + \beta_2 X. \label{eq:gomp}
\end{equation}

@@

\begin{equation}
\log(\lambda(t; X)) = \beta_0 + \beta_1 \log(t) + \beta_2 X. \label{eq:weibull}
\end{equation}

@@

To perform variable selection on the regression parameters $\theta \in \mathbb{R}^p$ of the hazard function, we can add a penalty to the likelihood and optimise the following equation:
\begin{equation}
\min _{\theta \in \mathbb{R}^{p}}\,\,-\ell\left(\theta\right)+\sum_{j=1}^p w_j p_{\lambda,\alpha}(\theta_j) \label{eq:penest}
\end{equation}
where $\ell\left(\theta\right) = \log L(\theta)$ is the log of the likelihood function given in \eqref{eq:lik-function}, $p_{\lambda,\alpha}(\theta_j)$ is a penalty term controlled by the non-negative regularization parameters $\lambda$ and $\alpha$, and $w_j$ is the penalty factor for the $j$th covariate. These penalty factors serve as a way of allowing parameters to be penalized differently. For example, we could set the penalty factor for time to be 0 to ensure it is always included in the selected model.

@@

Finally, several packages include functions to estimate the survival function and the CI. The corresponding methods generally fall into two categories: transformation of the estimated hazard function, and semi-parametric estimation of the baseline hazard. The first category broadly corresponds to parametric survival models, where the full hazard is explicitly modeled. Using this estimate, the survival function and the CI can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CDF} below). Packages providing this functionality include \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, and \pkg{survival}. Our package \pkg{casebase} also follows this approach for both single-event and competing risk analyses. The second category outlined above broadly corresponds to semi-parametric models. These models do not model the full hazard function, and therefore the baseline hazard needs to be estimated separately in order to estimate the survival function. This is achieved using semi-parametric estimators (e.g. Breslow's estimator) or parametric estimators (e.g. spline functions). Packages that implement this approach include \pkg{riskRegression}, \pkg{rstpm2}, \pkg{survival}, and \pkg{glmnet}. As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the survival function, whereas the second category often produces estimates in the form of stepwise functions.

@@

Once a model-fit object has been returned by `fitSmoothHazard`, all the familiar summary and diagnostic functions are available: `print`, `summary`, `predict`, `plot`, etc. Our package provides one more functionality: it computes risk functions from the model fit. For the case of a single event, it uses the familiar identity
\begin{equation}\label{eqn:surv}
S(t) = \exp\left(-\int_0^t \lambda(u;X) du\right).
\end{equation}
The integral is computed using either the numerical or Monte-Carlo integration. The risk function (or cumulative distribution function) is then defined as
\begin{equation}\label{eqn:CDF}
F(t) = 1 - S(t).
\end{equation}

@@

axis(2, at = seq(0,1.50, by=0.25), las = 2)
axis(1, at = seq(min(new_time),max(new_time), by=1))
for( i in seq(0,1.50, by=0.25)) {
  abline(h = i, lty = 1, col = "lightgrey")
}
for( i in seq(min(new_time),max(new_time), by=1)) {
  abline(v = i, lty = 1, col = "lightgrey")
}
for(i in seq_along(seq(0, 0.75, by = 0.25))) {
  mtext(paste0(seq(0, 0.75, by = 0.25)[i]*100,"%"), side = 4, las = 2, adj = 1, outer = TRUE, line = 2, at = c(0.63-0.11*(i-1)))
}
lines(tt$Follow.Up.Time, tt$hazard_ratio, lwd = 2, lty = 1)
mtext("% reduction in\nprostate cancer\nmortality rate", side = 4, las = 2, outer = TRUE, line = -1, at = 0.7)
```

@@

```{r, echo = TRUE, eval = FALSE}
fitSmoothHazard(DeadOfPrCa ~ as.integer(Follow.Up.Time >= t0) :
                    ScrArm : I(Follow.Up.Time - t0),
                data = ERSPC)
```

\noindent Here, we use the binary variable `as.integer(Follow.Up.Time >= t0)` in order to write the two cases of our formula above in a more compact way. We also use the function `I()`, which allows us to specify new variables within the formula using normal `R` code. This is required, as otherwise the symbol `-` would be interpreted as a formula operator. Finally, the term `ScrArm : I(Follow.Up.Time - t0)` represents the product $Z (t - t_0)$ in the equation above. We use the operator `:` instead of `*` in order to only include the interaction term, not the main effects.

@@

For the third case study, we show how \pkg{casebase} can also be used for variable selection through regularized estimation of the hazard function as given by Equation \ref{eq:penest}. We note that this is different than the semiparametric model Coxnet, which regularizes the Cox partial likelihood. To illustrate this functionality, we use the dataset from the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) [@knaus1995support].^[The original data is available online from the Department of Biostatistics at Vanderbilt University: https://biostat.app.vumc.org/wiki/Main/SupportDesc] The SUPPORT dataset tracks death in five American hospitals within individuals who are considered seriously ill. The cleaned and imputed data consists of 9104 observations and 30 variables, and it is available as part of the \pkg{casebase} package. In the comparisons below, all covariates except `sps` and `aps` were used. These two variables correspond to scores for predicting the outcome that were developed as part of the original study. For more information about this dataset, the reader is encouraged to look at the documentation in our package.

For our penalized case-base model, we opt for the natural log of time which corresponds to a Weibull distribution. For fitting the penalized hazard, we use `fitSmoothHazard.fit`, which is a matrix interface to the `fitSmoothHazard` function. The `fitSmoothHazard` and `fitSmoothHazard.fit` functions sample the case and base series, calculate the required offset, and transform the data to match the expected input of the \pkg{glmnet} package. The penalized logistic regression is then fit for multiple values of the tuning parameter using the function `glmnet::cv.glmnet` and the binomial family. To `fitSmoothHazard.fit`, we supply both a matrix `y` containing the time and event variables, and a matrix `x` containing all other covariates. We apply the lasso penalty by setting `alpha = 1` and assign a `penalty.factor` ($w_j$; cf. Equation \ref{eq:penest}) of 0 to the time variable to ensure it is in the selected model. We compare our approach to both Cox regression, and lasso penalized Cox regression (fitted via the \pkg{glmnet} package and using the Cox family).

>>>>

Case-base sampling allows us to avoid this integral. By conditioning on a sampled person-moment, we get individual partial likelihood contributions of the form
$$P(\,\textrm dN_{i}(t) \mid \,\textrm dQ_{i}(t) = 1,\mathrm{past}) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{\,\textrm dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.$$
By using the symbol $\stackrel{\theta}{\propto}$, we mean that both sides of the expression are equal up to multiplicative factors that do not depend on $\theta$. Therefore, an estimating function for $\theta$ can be composed of these contributions as:
\begin{equation}
L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{\,\textrm dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{\,\textrm dQ_i(t)}. (\#eq:lik-function)
\end{equation}
When a logarithmic link function is used for modeling the hazard function, the above expression is of a logistic regression form with an offset term $\log(1/\rho_i(t))$. Note that the sampling units selected in the case-base sampling mechanism are person-moments, rather than individuals, and the parameters to be estimated are hazards or hazard ratios rather than odds or odds ratios. Generally, an individual can contribute more than one person-moment, and thus the terms in the product integral are not independent. Nonetheless, Saarela [-@saarela2016case] showed that the corresponding partial likelihood score function has mean zero at the true value $\theta=\theta_0$, and that the resulting estimator $\widehat{\theta}$ is asymptotically normally distributed.

@@

  1. Compute $\widehat{\theta}$ and its variance-covariance matrix $V(\widehat \theta)$ which is obtained as part of the logistic regression output.
  2. Sample $B$ times from a multivariate normal $N\left(\widehat{\theta}, V(\widehat\theta)\right)$.
  3. For each sample, compute the survival function using Equation \@ref(eq:surv) below.
  4. Use the pointwise quantiles of these survival function estimates to construct a pointwise confidence band for the survival function of interest.

@@

\begin{equation}
\log(\lambda(t; X)) = \beta_0 + \beta_1 X. (\#eq:exp)
\end{equation}
In this model, the instantaneous failure rate is independent of $t$.^[The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study. This is also known as the *memoryless property* [@kalbfleisch2011statistical].]

@@

\begin{equation}
\log(\lambda(t; X)) = \beta_0 + \beta_1 t + \beta_2 X. (\#eq:gomp)
\end{equation}

@@

\begin{equation}
\log(\lambda(t; X)) = \beta_0 + \beta_1 \log(t) + \beta_2 X. (\#eq:weibull)
\end{equation}

@@

To perform variable selection on the regression parameters $\theta \in \mathbb{R}^p$ of the hazard function, we can add a penalty to the likelihood and optimise the following equation:
\begin{equation}
\min _{\theta \in \mathbb{R}^{p}}\,\,-\ell\left(\theta\right)+\sum_{j=1}^p w_j p_{\lambda,\alpha}(\theta_j) (\#eq:penest)
\end{equation}
where $\ell\left(\theta\right) = \log L(\theta)$ is the log of the likelihood function given in \@ref(eq:lik-function), $p_{\lambda,\alpha}(\theta_j)$ is a penalty term controlled by the non-negative regularization parameters $\lambda$ and $\alpha$, and $w_j$ is the penalty factor for the $j$th covariate. These penalty factors serve as a way of allowing parameters to be penalized differently. For example, we could set the penalty factor for time to be 0 to ensure it is always included in the selected model.

@@

Finally, several packages include functions to estimate the survival function and the CI. The corresponding methods generally fall into two categories: transformation of the estimated hazard function, and semi-parametric estimation of the baseline hazard. The first category broadly corresponds to parametric survival models, where the full hazard is explicitly modeled. Using this estimate, the survival function and the CI can be obtained using their functional relationships (see Equations \@ref(eq:surv) and \@ref(eq:CDF) below). Packages providing this functionality include \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, and \pkg{survival}. Our package \pkg{casebase} also follows this approach for both single-event and competing risk analyses. The second category outlined above broadly corresponds to semi-parametric models. These models do not model the full hazard function, and therefore the baseline hazard needs to be estimated separately in order to estimate the survival function. This is achieved using semi-parametric estimators (e.g. Breslow's estimator) or parametric estimators (e.g. spline functions). Packages that implement this approach include \pkg{riskRegression}, \pkg{rstpm2}, \pkg{survival}, and \pkg{glmnet}. As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the survival function, whereas the second category often produces estimates in the form of stepwise functions.

@@

Once a model-fit object has been returned by `fitSmoothHazard`, all the familiar summary and diagnostic functions are available: `print`, `summary`, `predict`, `plot`, etc. Our package provides one more functionality: it computes risk functions from the model fit. For the case of a single event, it uses the familiar identity
\begin{equation}
S(t) = \exp\left(-\int_0^t \lambda(u;X) du\right). 
(\#eq:surv)
\end{equation}
The integral is computed using either the numerical or Monte-Carlo integration. The risk function (or cumulative distribution function) is then defined as
\begin{equation}
F(t) = 1 - S(t). (\#eq:CDF)
\end{equation}

@@

axis(2, at = seq(0,1.50, by = 0.25), las = 2)
axis(1, at = seq(min(new_time),max(new_time), by = 1))
for (i in seq(0,1.50, by = 0.25)) {
  abline(h = i, lty = 1, col = "lightgrey")
}
for (i in seq(min(new_time),max(new_time), by = 1)) {
  abline(v = i, lty = 1, col = "lightgrey")
}
for (i in seq_along(seq(0, 0.75, by = 0.25))) {
  mtext(paste0(seq(0, 0.75, by = 0.25)[i]*100,"%"), side = 4, las = 2, adj = 1, outer = TRUE, line = 2, at = c(0.63 - 0.11*(i - 1)))
}
lines(tt$Follow.Up.Time, tt$hazard_ratio, lwd = 2, lty = 1)
mtext("% reduction in\nprostate cancer\nmortality rate", side = 4, las = 2, outer = TRUE, line = -1, at = 0.7)
```

@@

```{r, echo = TRUE, eval = FALSE}
fitSmoothHazard(DeadOfPrCa ~ as.integer(Follow.Up.Time >= t0) :
                    ScrArm : I(Follow.Up.Time - t0),
                data = ERSPC)
```
Here, we use the binary variable `as.integer(Follow.Up.Time >= t0)` in order to write the two cases of our formula above in a more compact way. We also use the function `I()`, which allows us to specify new variables within the formula using normal `R` code. This is required, as otherwise the symbol `-` would be interpreted as a formula operator. Finally, the term `ScrArm : I(Follow.Up.Time - t0)` represents the product $Z (t - t_0)$ in the equation above. We use the operator `:` instead of `*` in order to only include the interaction term, not the main effects.

@@

For the third case study, we show how \pkg{casebase} can also be used for variable selection through regularized estimation of the hazard function as given by Equation \@ref(eq:penest). We note that this is different than the semiparametric model Coxnet, which regularizes the Cox partial likelihood. To illustrate this functionality, we use the dataset from the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) [@knaus1995support].^[The original data is available online from the Department of Biostatistics at Vanderbilt University: https://biostat.app.vumc.org/wiki/Main/SupportDesc] The SUPPORT dataset tracks death in five American hospitals within individuals who are considered seriously ill. The cleaned and imputed data consists of 9104 observations and 30 variables, and it is available as part of the \pkg{casebase} package. In the comparisons below, all covariates except `sps` and `aps` were used. These two variables correspond to scores for predicting the outcome that were developed as part of the original study. For more information about this dataset, the reader is encouraged to look at the documentation in our package.

For our penalized case-base model, we opt for the natural log of time which corresponds to a Weibull distribution. For fitting the penalized hazard, we use `fitSmoothHazard.fit`, which is a matrix interface to the `fitSmoothHazard` function. The `fitSmoothHazard` and `fitSmoothHazard.fit` functions sample the case and base series, calculate the required offset, and transform the data to match the expected input of the \pkg{glmnet} package. The penalized logistic regression is then fit for multiple values of the tuning parameter using the function `glmnet::cv.glmnet` and the binomial family. To `fitSmoothHazard.fit`, we supply both a matrix `y` containing the time and event variables, and a matrix `x` containing all other covariates. We apply the lasso penalty by setting `alpha = 1` and assign a `penalty.factor` ($w_j$; cf. Equation \@ref(eq:penest)) of 0 to the time variable to ensure it is in the selected model. We compare our approach to both Cox regression, and lasso penalized Cox regression (fitted via the \pkg{glmnet} package and using the Cox family).
